---
title: "Curating an eDNA Reference Database for Slovenian Freshwater Fishes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Curating a DNA Reference Database with barcurateR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>", eval=FALSE)
library(barcurateR)
library(dplyr)
library(readr)
library(tidyr)
library(Biostrings)
library(RefManageR)

BibOptions(check.entries = FALSE, style = "html", hyperlink = FALSE, cite.style = 'authoryear', max.names = 1, longnamesfirst = F)
myBib <- ReadBib(system.file("extdata", "barcurateR.bib", package = "barcurateR"), check = FALSE)

outputs <- readRDS(system.file("extdata", "case_study_outputs.rds", package = "barcurateR"))
final_assignments <- read_csv(system.file("extdata", "sava_river_12s_assignments_two_pass.csv", package = "barcurateR"))
community_matrix <- read_csv(system.file("extdata", "sava_river_community_matrix.csv", package = "barcurateR"))
```

# Introduction

`barcurateR` provides a reproducible workflow for building, auditing, and exporting curated DNA barcode reference databases for eDNA metabarcoding, which includes basic taxonomic assignment functions, i.e., BLASTn and exact matching. 

This vignette walks through a realistic freshwater fish case study: curating a local reference database using NCBI and BOLD sequences to be used for the taxonomic assignment for a eDNA dataset.

## Dataset

For this example, we used the Slovenian fish species list presented by the Slovenian Native Fish Society `r Cite(myBib, 'drustvo_za_preucevanje_rib_slovenije_nase_2018')`, supplemented by "The Updated List of Non-Native Freshwater Fishes in Slovenia with Note of their Potential Impacts in Inland Waters" `r Cite(myBib, 'povz_updated_2018')`, for a total of 97 species. The sequences from NCBI (queried and retrieved on 2026-10-02) and BOLD (retrieved on 2026-09-16) were parsed and subsetted separately based on the compiled species list.

The reference database validation was conducted on fish eDNA dataset from the middle reach of the Sava River as reported from the study: "Temporal and Spatial Dynamics of the Lotic Fish Communities: A Comparison of Coffee Filter-Based Passive eDNA Collection Versus Active eDNA Filtering" `r Cite(myBib, 'mlinarec_temporal_2025')`. The validation eDNA ASV was reconstructed following the DADA2 workflow `r Cite(myBib, 'callahan_dada2_2016')` using parameters described by `r Cite(myBib, 'mlinarec_temporal_2025', textual=TRUE)`, where applicable.

*Note: File names and paths may need to be replaced when processing your own data; curation and taxonomic assignment steps may require external software i.e., BLAST, MAFFT, and FastTree.*

# 1. Setup and Configuration

## Loading libraries

`dplyr`, `reader`, and `tidyr` are used for data manipulation and `Biostrings` for FASTA handling.

```{r message=FALSE, warning=FALSE}
library(barcurateR)
library(dplyr)
library(readr)
library(tidyr)
library(Biostrings)
```

## Loading the Dataset

For this case study, reference sequences retrieved from NCBI (`ncbi_raw`) and BOLD (`bold_raw`), along with a taxonomic table (`taxonomy_table`) for the target species list.

```{r}
ncbi_csv <- "data-raw/slovenian_fish/raw/ncbi_sequence_raw.csv"
bold_csv <- "data-raw/slovenian_fish/raw/bold_sequence_raw.csv"
taxonomy_csv <- "data-raw/slovenian_fish/processed/slovenian_fish_taxonomy.csv"

ncbi_raw <- read_csv(ncbi_csv, show_col_types = FALSE) |>
  filter(!is.na(sequence))

bold_raw <- read_csv(bold_csv, show_col_types = FALSE) |>
  filter(!is.na(sequence))

taxonomy_table <- read_csv(taxonomy_csv, show_col_types = FALSE)
```

# 2. Standardization & ambiguity resolution

## Data standardization

We use `rb_parse_source_table()` to standardize the heterogeneous inputs. This function extracts sequence marker information from the sequence headers (common in NCBI/BOLD downloads). `rb_parse_source_table()` allows users to provide the column mapping and extracts additional marker information from the `description_col` as typically included in the header from NCBI and BOLD data. Detailed descriptions for required variables are described in the manual for `rb_parse_source_table()`.

```{r}
# NCBI
ncbi_raw$sequence_id <- sub("\\s.*", "", ncbi_raw$header)

ncbi_parsed <- rb_parse_source_table(
  raw = ncbi_raw,
  source_name = "ncbi",
  column_map = c(
    sequence_id = "sequence_id",
    species = "species",
    sequence = "sequence"
  ),
  description_col = "header"
)

# BOLD
bold_raw$description <- bold_raw$seq_type

bold_parsed <- rb_parse_source_table(
  raw = bold_raw,
  source_name = "bold",
  column_map = c(
    sequence_id = "sequence_id",
    species = "species_name",
    sequence = "sequence"
  ),
  description_col = "description"
)
```

## Data merge

`rb_combine_source()` automatically merges standardized sequence data and deduplicates identical species-sequence entries.

```{r}
combined <- rb_combine_sources(list(ncbi_parsed, bold_parsed))
```

## Ambiguity resolution

Often, sequences downloaded from large public repositories, e.g., NCBI, also contain identical sequences attributed to different taxa. `rb_detect_ambiguity()` helps users identify these sequences by flagging many-to-one species-sequence data for subsequent action and exports the ambiguities as a CSV as documentation. Users can provide an ambiguity resolution table for `rb_resolve_ambiguous()`, confirm that the sequence dataset no longer has any ambiguities (`"stop"` or `"warn"`), or as we show here, to automatically drop all ambiguous sequences.

```{r}
combined <- rb_detect_ambiguity(combined)
n_ambig <- sum(combined$is_ambiguous)

if (n_ambig > 0) {
  rb_write_ambiguity_csv(
    combined[combined$is_ambiguous, ],
    "data-raw/slovenian_fish/processed/ambiguous_sequences.csv"
  )
}

resolved <- rb_resolve_ambiguous(
  combined,
  on_unresolved = "drop"
)
```

# 3. Curation Pipeline

The curation workflow for `barcurateR` reproduces the pipeline described from "YZFishDB: A curated DNA reference database for the fishes of the Yangtze River Basin" `r Cite(myBib, 'kua_yzfishb_2026')`. the core function `rb_curate_reference()` completes the entire curation workflow as described by `r Cite(myBib, 'kua_yzfishdb_2026', textual=TRUE)` and exports as SQLite database, while allowing users to opt for major steps in the workflow. Specific requirements and dependencies for `rb_curate_reference()` are detailed in the manual (`?rb_curate_reference()`). 

For this case study, we passed our combined sequence dataset through the full curation workflow by providing the contaminant `BLASTn` database and the numts `FASTA`. 

*Note: External software BLAST+ (`blastn` and `makeblastdb`), MAFFT, and FastTree are required to complete the full curation pipeline*

```{r}
contam_path <- "tests/testthat/fixtures/contaminants/contaminant_db"
numts_path  <- "tests/testthat/fixtures/contaminants/fish_numts.fasta"
db_path <- "data-raw/slovenian_fish/processed/slovenian_fish_refdb.sqlite"

curated <- rb_curate_reference(
  data = resolved,
  blast_db = contam_path,
  numt_fasta = numts_path,
  run_classifier = TRUE,
  run_divergence = TRUE,  # requires MAFFT and FastTree
  run_barcode_gap = TRUE,
  taxonomy_table = taxonomy_table,
  db_path = db_path,
  on_ambiguous = "stop",
  require_taxonomy = TRUE,
  archive_ambiguity = TRUE,
  ambiguity_path = "data-raw/slovenian_fish/processed/ambiguous_sequences.csv"
)
```

# 4. Database auditing

The curated reference database can be queried and audited through several built-in functions to inspect and summarize information e.g., taxonomic and marker coverage. These functions allow users to better understand the composition, quality, and coverage of the reference database. Curated `barcurateR` reference database can be connected through `rb_connect()`. 

```{r}
con <- rb_connect(db_path)
```

## Reference database sequence quality and coverage

### Audit QC flags

```{r}
qc_sum <- rb_qc_summary(con)
print(qc_sum)
```

```{r eval=TRUE}
print(outputs[[1]])
```

### Audit marker coverage

```{r}
marker_cov <- rb_marker_coverage(con, rank = "order")
head(marker_cov)
```

```{r eval=TRUE}
head(outputs[[2]])
```

### Audit taxonomic coverage

```{r}
taxa_cov <- rb_list_taxa(con, rank = "species")
head(taxa_cov)
```

```{r eval=TRUE}
head(outputs[[3]])
```

## Barcode gap analysis

To help users evaluate the efficacy of taxonomic assignments, `barcurateR` includes a simple barcode gap analysis.

```{r}
gap_metrics <- rb_barcode_gap(con)

if (nrow(gap_metrics) > 0) {
  poor_resolution <- gap_metrics |>
    filter(taxonomic_resolution != "species" | assignment_risk == "high")
}

head(poor_resolution)
```

```{r eval=TRUE}
head(outputs[[4]])
```

```{r}
rb_disconnect(con)
```

# 5. Downstream Applications

## Exporting database

For users conducting most of the bioinformatics and downstream taxonomic assignment work in other popular platforms, `barcurateR` provides a simple wrapper function to export the selected reference database for i) DADA2, ii) QIIME2, iii) BLASTn, iv) USEARCH/VSEARCH. Here we first subset the curated database based on the `marker` and `qc_flag` according to our needs using `rb_get_sequences()` before exporting them. 

```{r}
con <- rb_connect(db_path)

refs_12S <- rb_get_sequences(con, marker = "12S", qc_flag = "pass")
refs_COI <- rb_get_sequences(con, marker = "COI", qc_flag = "pass")

# DADA2
rb_export_dada2(
  refs_12S,
  "exports/slovenian_fish_12S_dada2.fasta"
)

# QIIME2
rb_export_qiime2(
  refs_COI,
  "exports/slovenian_fish_COI_qiime2.fasta",
  "exports/slovenian_fish_COI_taxonomy.tsv"
)

# BLASTn: 12S only
rb_export_blastn(
  rb_get_sequences(con, marker = "12S", qc_flag = "pass"),
  "exports/slovenian_fish_12s_blastn.fasta"
)

# BLASTn: all markers
rb_export_blastn(
  rb_get_sequences(con, qc_flag = "pass"),
  "exports/slovenian_fish_all_markers_blastn.fasta"
)

rb_disconnect(con)
```

## Taxonomic assignment

`barcurateR` includes two taxonomic assignment methods, i.e., BLASTn and exact matching. BLASTn taxonomic assignment relies on the external software from the BLAST+ suite (`blastn`) while exact matching is a dependency-free method that was included for testing and quick assignment jobs. 

The package requires ASV sequences in a FASTA file (`sava_asvs_fasta`) and a corresponding count table (`sava_counts`).

```{r}
sava_asvs_fasta <- "data-raw/sava_river/processed/sava_asvs.fasta"
sava_counts <- read_csv(
  "data-raw/sava_river/processed/sava_abundance_table.csv",
  show_col_types = FALSE
)
```

For this example, we conducted a two-pass taxonomic assignment using BLASTn: 

  Pass 1: Assign against the 12S subset of the curated reference database (`marker="12S"` and `qc_flag="pass"`).
  Pass 2: Assign unassigned ASVs against the full reference database (`marker=NULL` and `qc_flag=c("pass","manual_review_needed")`) to rescue borderline matches.

```{r}
con <- rb_connect(db_path)

# Pass 1: 12S subset
res_pass1 <- rb_assign_edna(
  asv_fasta = sava_asvs_fasta,
  con = con,
  marker = "12S",
  method = "blastn",
  min_identity = 99,
  out_dir = "exports/blastn_pass1"
)

all_asv_ids <- names(Biostrings::readDNAStringSet(sava_asvs_fasta))
assigned_ids <- res_pass1$asv_id[res_pass1$assignment_status != "no_match"]
unassigned_ids <- setdiff(all_asv_ids, assigned_ids)

# Pass 2: full reference for unassigned ASVs
if (length(unassigned_ids) > 0) {
  all_seqs <- Biostrings::readDNAStringSet(sava_asvs_fasta)
  pass2_fasta <- tempfile(fileext = ".fasta")

  Biostrings::writeXStringSet(
    all_seqs[names(all_seqs) %in% unassigned_ids],
    pass2_fasta
  )

  res_pass2 <- rb_assign_edna(
    asv_fasta = pass2_fasta,
    con = con,
    marker = NULL,
    method = "blastn",
    qc_flag = c("pass", "manual_review_needed"),
    min_identity = 99,
    min_coverage = 0.9,
    max_target_seqs = 500,
    out_dir = "exports/blastn_pass2"
  )

  res_pass2$db_source <- "full_reference"
} else {
  res_pass2 <- res_pass1[0, ]
}

rb_disconnect(con)

# combine both assignment tabls
final_assignments <- bind_rows(
  res_pass1 |>
    filter(assignment_status != "no_match") |>
    mutate(db_source = "12S_subset"),
  res_pass2
)

missing_ids <- setdiff(all_asv_ids, final_assignments$asv_id)

if (length(missing_ids) > 0) {
  pad <- final_assignments[rep(NA_integer_, length(missing_ids)), , drop = FALSE]
  pad$asv_id <- missing_ids
  pad$assignment_status <- "no_match"
  pad$db_source <- "no_hit"
  final_assignments <- bind_rows(final_assignments, pad)
}

head(final_assignments)

# export taxonomic assignment table
write_csv(
  final_assignments,
  "exports/sava_river_12s_assignments_two_pass.csv"
)
```

```{r eval=TRUE}
head(final_assignments)
```

After the taxonomic assignment is completed, we can use `rb_build_species_matrix()` to create the species-site data table using the taxonomic assignment (`final_assignments`) and the count table (`sava_counts`).

```{r}
community_matrix <- rb_build_species_matrix(
  assignment = final_assignments,
  asv_table = sava_counts,
  include_ambiguous = FALSE,
  unassigned = FALSE
)

community_matrix[1:5, 1:5]

# export species-site matrix
write_csv(
  community_matrix,
  "exports/sava_river_community_matrix.csv"
)
```

```{r eval=TRUE}
community_matrix[1:5, 1:5]
```

# Conclusion

This case study demonstrates how `barcurateR` streamlines the complex process of reference database curation. By packaging robust QC, ambiguity resolution, and auditing tools, `barcurateR` ensures that downstream eDNA assignments are based on high-quality, transparent, and reproducible reference data.

# References

```{r eval=TRUE}
PrintBibliography(myBib)
```
